---
title: "Advanced Statistics"
author: "Zoe Schroder"
date: "9/22/2019"
output: html_document
---

```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(msme))
suppressMessages(library(lme4))
```


*Distributions: *

A distribution is a function that shows the possible values for a variable and how often they occur. It is considered the shape of the data and represents the counts of the data that fall in that column/bar graph.  

```{r}
load("BigDays.RData")
```

A discrete distribution is a statistical distribution that shows the probabilities of outcomes with finite values. It describes the probability of occurrence for discrete events. Discrete values can only take particular values. There may be infinite nubme rof those values, but each is distinct. It can be numeric like counts of things or categorical like yes or no. Whole numbers (integers) are discrete variables.  

```{r}
ggplot(BigDays.sfdfT, aes(x = Mo)) +
  geom_bar()
```

A continuous distribution is built from outcomes that potentially have infinite measurable values. These data are not restricted to separate values. These values are always numeric. (i.e. Decimals)

```{r}
ggplot(BigDays.sfdfT, aes(x = avgCAPE)) +
  geom_bar(binwidth = 50)
```

`Poisson distribution` is a discrete probability distribution. It expresses the probability of a number of events occurring within a specified period of time. It is used for independent events that occur over a constant period of time. The poisson distribution has one parameter $lambda$ which represents the mean and variance of the model. A variable follows a poisson distribution if: 

1. Data are counts of events (only positive; no upper bound)
2. Events are independent
3. Average rate does not change over the time period of interest. 

```{r}
pois_big <- as.data.frame(rpois(n = 1000, lambda = 50))
names(pois_big) <- c("counts")

ggplot(pois_big, aes(x = counts)) +
  geom_bar()

pois_small <- as.data.frame(rpois(n = 1000, lambda = 1))
names(pois_small) <- c("counts")

ggplot(pois_small, aes(x = counts)) +
  geom_bar(binwidth = 1)
```


The number of tornadoes follows a poisson distribution. 
```{r}
ggplot(BigDays.sfdfT, aes(x = nT)) +
  geom_bar(binwidth = 1)
```

The `Binomial Distribution` is a discrete distribution that models the number of events in a fixed number of trials. Each trial has two possible outcomes and event is the outcome of interest from the trial. The binomial distribution is used when you are interested in the occurrence of an event and not its magnitude. It sets an upper limit on the count as the number of events you observe cannot be greater than the number of trials you perform. It is a binomial distribution if: 

1. The number of trials is fixed
2. Each trial is independent of the other trials
3. Each trial has one of two outcomes: event or nonevent
4. The probability of an event is the same for each trial. 

When the number of trials is large, the binomial distribution can be reasonably approximated by the normal distribution. 

n = 100, p = 0.5

```{r}
binomial <- as.data.frame(rbinom(n = 1000, size = 50, prob = 0.5), col.names = count)
names(binomial) <- c("counts")

ggplot(binomial, aes(x = counts)) +
  geom_bar(binwidth = 1)
```

The `Lognormal Distribution` is used when the logarithm of the random variable is normally distributed. This is used only when the variables are positive values. It is a continuous distribution that is defined by its location and scale parameters. In total, it is characterized by 3 paramters of its location, scale, and threshold parameters. 

```{r}
lognormal <- as.data.frame(rlnorm(n = 1000, meanlog = 0, sdlog = 1))
names(lognormal) <- c("counts")

ggplot(lognormal, aes(x = counts)) +
  geom_bar(binwidth = 0.5)

ggplot(lognormal, aes(x = log(counts))) +
  geom_bar(binwidth = 0.5)
```


Accumulate Tornado Power on big tornado days is an example of the lognormal distribution. 
```{r}
ggplot(BigDays.sfdfT, aes(x = ATP_TW)) +
  geom_bar(binwidth = 2)

ggplot(BigDays.sfdfT, aes(x = log(ATP_TW))) +
  geom_bar(binwidth = 0.25)
```

The `Exponential distribution` can be used to model the time between events in a continuous Poisson process. The events are independent and occur at a constant rate. It has two parameters: scale and threshold ($theta$). For a positive threshold parameter, the distribution is shifted to the right by a distance $theta$. 

```{r}
exponential <- as.data.frame(rexp(n = 1000, rate = 10))
names(exponential) <- c("counts")

ggplot(exponential, aes(x = counts)) +
  geom_bar(binwidth = 0.01)
```

A `Normal Distribution` is a continuous distribution that is specified by the mean ($mu$) and the standard deviation ($sigma$). The mean is the peak or center of the bell-shaped curve and the standard deviation determines the spread of the distribution. Normal distributions are the most common because many things occur naturally in physical, biological, and social measurement situations. 

```{r}
normal <- as.data.frame(rnorm(n = 1000, mean = 0, sd = 1))
names(normal) <- c("counts")

ggplot(normal, aes(x = counts)) +
  geom_bar(binwidth = 0.1)

```

A `t-distribution` is used to analyze the mean of an approximately normal population when the population standard deviation is unknown. You can use it to test if the population mean is different from the hypothesized mean. It is a  continuous distribution that is specified by the number of degrees of freedom. It is bell-shaped and similar to the normal distribution however it has thicker tails. 

```{r}
tdist <- as.data.frame(rt(n = 1000, df = 100))
names(tdist) <- c("counts")

ggplot(tdist, aes(x = counts)) +
  geom_bar(binwidth = .1)

```

The `Gamma Distribution` is used to model continuous variables that are always positive and have skewed distributions. It has two parameters: shape ($alpha$) and scale ($beta$). It is related to the normal and exponential distributions. 

```{r}
gam <- as.data.frame(rgamma(n = 1000, shape = 2, rate = 4))
names(gam) <- c("counts")

ggplot(gam, aes(x = counts)) +
  geom_bar(binwidth = 0.1)
```

The `Beta Distribution` is a family of continuous probability distributions parameterized by two positive shape paramters that appear as exponents of the random variable and control the shape of the distribution. It is described as a probability distribution of probabilities. It represents all possible values of a probability when we don't know what that probability is. Think of batting averages, the more you have the lower the average becomes and the more narrow the data fluctuation becomes. 

```{r}
beta <- as.data.frame(rbeta(n = 1000, shape1 = 3, shape2 = 5))
#shape1: as it increases it shifts the peak to the right. 
#shape2: adjusts the count range on the y-axis
names(beta) <- c("counts")

ggplot(beta, aes(x = counts)) +
  geom_bar(binwidth = 0.01)
```

The `Negative Binomial Distribution` is a discrete probability distribution counts the number of failures until a fixed number of successes. It is a draw with replacement. It is characterized by: 

1. experiment consists of x repeated trials
each trial has two possible outcomes
3. The probability of success is the same in every trial
4. The trials are inndependent
5. The experiment continues until a certain number (x) of successes are observed

Has two parameters: lambda and one that adjusts the variance independently from the mean (corrects overdispersion). 

```{r}
negbinom <- as.data.frame(rnbinom(n = 1000, size = 1, mu = 5))
names(negbinom) <- c("counts")

ggplot(negbinom, aes(x = counts)) +
  geom_bar(binwidth = 1)
```






To check for overdispersion, calculate the Pearson Chi-squared statistic on the fitted model. You should also plot the standardized deviance residuals. 
```{r}
BigDays.sfdfT$maxCAPEs <- scale(BigDays.sfdfT$maxCAPE)
BigDays.sfdfT$maxBSs <- scale(BigDays.sfdfT$maxBS)

model <- glm(nT ~ maxCAPEs + maxBSs, 
                    data = BigDays.sfdfT, 
                    family = "poisson")

summary(model)

#Values greater than 1 indicate model over dispersion. Less than 1 indicate under-dispersion
P__disp(model) 

plot(rstandard(model))
#a good fitting model will have the points between -2 and 2

model <- glmer(nT ~ maxCAPEs + maxBSs + (1 | Mo), 
                    data = BigDays.sfdfT, 
                    family = poisson)

# overdispersion = deviance / residual degrees of freedom
summary(model)$sigma
deviance(model) / df.residual(model)

overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}
overdisp_fun(model)
```

*Linear Mixed Effects Models*

A mixed model is a linear model that includes both `Fixed` and `Random` effects in the same model. A fixed effect is a measure of the population level with the assumption that the $beta$ coefficient for each fixed effect is the same for all subjects. Fixed effects can be continuous or categorical. Variables are fixed when no matter where, how, or how many are sampled, you would have the same levels of the same variables. A random effect is not constant between all subjects and is a measure at the group level. It is an estimate of the variance of the effect. Random effects are categorical. Random effects do not remain fixed between samples and would change depending on increased sampling or a change in how the data was collected. Random effects can be nested or crossed. Nested random effects are when the each effect is impacted by the other (i.e. the observer assigns the test id). Crossed is when the random effect is not influenced by something else (i.e. genetic lineages from wasps). The syntax for a mixed model is: 
$$
Y = \alpha X +\beta Z + \epsilon
$$
where Y is the vector of observed values, $alpha$ is a vector of fixed parameters, $beta$ is a vector of random effect parameters ($B ~ N(0,G)$), $X$ is the Fixed Effects Design matrix, $Z$ is the Random Effects Design matrix, and $epsilon$ is the vector of residuals. 

You should use a mixed model instead of a simple linear model when you have a viariable that describes your data sample as a subset as a subset of the data you have collected. Mixed models are used in situations where: 
1. Incomplete information
2. In cases with fewer treatment periods than treatments
3. Analysis of multi-center trial
4. Repeated measurements over time
5. Familial data sets

How do you find a probability distribution that best fits your data? 
```{r}
library(car)
library(MASS)
```

Check out the plots generated using qqp. The y axis represents the observations and the x axis represents the quantiles modeled by the distribution. The solid red line represents a perfect distribution fit and the dashed red lines are the confidence intervals of the perfect distribution fit. You want to pick the distribution for which the largest number of observations falls between the dashed lines. In this case, that's the lognormal distribution, in which only one observation falls outside the dashed lines. Now, armed with the knowledge of which probability distribution fits best, I can try fitting a model.

*NOTE: The negative binomial and gamma distributions can only handle positive numbers. The poisson distribution can only handle positive whole numbers. The binomial and poisson distributions are different from the others because they are discrete rather than continuous (they quantify distinct countable events or the probability of these events.)*

`Normal`
```{r}
qqp(BigDays.sfdfT$nT, "norm")
```

`Log-Normal`
```{r}
qqp(BigDays.sfdfT$nT, "lnorm")
```

```{r}
nbinom <- fitdistr(BigDays.sfdfT$nT, "Negative Binomial")
qqp(BigDays.sfdfT$nT, "nbinom", size = nbinom$estimate[[1]], mu = nbinom$estimate[[2]])
```

`Poisson`
```{r}
pois <- fitdistr(BigDays.sfdfT$nT, "Poisson")
qqp(BigDays.sfdfT$nT, "pois", lambda = pois$estimate)
```

`Gamma`
```{r}
gamma <- fitdistr(BigDays.sfdfT$nT, "gamma")
qqp(BigDays.sfdfT$nT, "gamma", shape = gamma$estimate[[1]], rate = gamma$estimate[[2]])
```

```{r}
BigDays.sfdfT$maxCAPEs <- scale(BigDays.sfdfT$maxCAPE)
BigDays.sfdfT$maxBSs <- scale(BigDays.sfdfT$maxBS)
BigDays.sfdfT$maxHLCYs <- scale(BigDays.sfdfT$maxHLCY)
BigDays.sfdfT$minCINs <- scale(BigDays.sfdfT$minCIN)

model <-  lmer(nT ~ maxCAPEs + maxBSs + maxHLCYs + minCINs + (1 | Mo),
               data = BigDays.sfdfT)
summary(model)
```

The estimate of the variance explained by the random effect needs to be distinguishable from 0 else the random effect has no effect on the model. To get p-values for your model: 

```{r}
Anova(model)
```


```{r}
model_pois <- glmer(nT ~ maxCAPEs + maxBSs + maxHLCYs + minCINs + (1 | Mo),
               data = BigDays.sfdfT,
               family = "poisson")
summary(model_pois)
```

Consider overdispersion...this occurs when the combined residuals are much larger than the residual degrees of freedom. 

```{r}
overdisp_fun <- function(model) {
    ## number of variance parameters in an n-by-n variance-covariance matrix
    vpars <- function(m) {
        nrow(m) * (nrow(m) + 1)/2
    }
    # The next two lines calculate the residual degrees of freedom
    model.df <- sum(sapply(VarCorr(model), vpars)) + length(fixef(model))
    rdf <- nrow(model.frame(model)) - model.df
    # extracts the Pearson residuals
    rp <- residuals(model, type = "pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    # Generates a p-value. If less than 0.05, the data are overdispersed.
    pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
    c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
}

overdisp_fun(model_pois)
```
*NOTE: You can model overdispersion as a random effect with one random effect level for each observation. As an example, using a unique ID for each event as a random effect would work.*













